### 2025-06-09 chapters/conclusion/conclusion.tex
Completed Dissertation Revision

Likelihood based inference for state space models is a large and challenging topic, and this thesis only presents a few innovations to solve this problem.
Below, I outline several future projects that build on the work presented in this dissertation.

@@

Additionally, the combination of MPIF and automatic differentiation may facilitate the use of shallow neural networks to be used in conjunction with mechanistic models.
As an example, consider a continuous time compartmental model outlined in Appendix~\ref{subsec:smcs}.
Transition rates between compartments are generally expressed as functions of latent and exogenous variables, denoted $f(\bm{X}(t), \bm{Z}(t))$.
Rather than predetermining the function $f$, automatic differentiation software may enable treating $f$ as a flexible function that needs to be estimated.
A similar idea has been explored previously \citep{noordijk24,dandekar20}, though only in the context of deterministic ODE models.

\subsection*{Applications in Ecology and Epidemiology}

The MPIF algorithm presented in Chapter~\ref{chpt:mpif} results in an improved capacity to calibrate PanelPOMP models to data via maximum likelihood.
Given the prevalence of panel data and applications of POMP models to univariate time series, this innovation may lead to a number of collaborative research projects in Ecology and Epidemiology.
Two application areas I feel are particularly promising are fisheries and livestock management.

Fisheries have a rich history of using state space models, though typically inference is performed using Lapacian approximations of the likelihood \citep[e.g.,][]{auger17}.
In addition to comparing whether this approach is a more reliable approximation that those built on sequential Monte Carlo methodologies, studying infectious diseases prevalent in native river and lake species may be uniquely well suited for PanelPOMP methodology: population dynamics in each river or lake are independent from other locations, yet share many similar features that should be estimated simultaneously.
Recently published research on livestock management \citep{skolstrup22} used linear Gaussian state space models to estimate the effects of interventions at dairy farms, and methodologies enabling nonlinear models may be a natural next step.

>>>>

Recent advances in machine learning and artificial intelligence (AI) have significantly enhanced researchers' ability to extract information from complex data sets.
Consequently, much of the recent enthusiasm and effort in statistics and computational science over the last decade has been dedicated to expanding the limits, capabilities, and applications of these fields.
While the excitement surrounding machine learning and AI is well-justified, state-space models (SSMs) also play a crucial role in advancing scientific understanding \citep{baker18,hogg24}.
For example, SSMs offer greater interpretability compared to machine learning algorithms, making them valuable for inferring features of dynamic systems.
Moreover, by incorporating real-world mechanisms into their formulations, SSMs have the potential to improve predictions and provide more reliable estimates of the impacts of system interventions.
This latter task is particularly challenging for machine learning algorithms, which depend on existing data, as hypothetical interventions represent counterfactual scenarios not present in the data.

Advances in SSM methodology may also play a crucial role in advancing machine learning and AI.
As summarized by \citet{lin24}, the large language models that have been the driving force behind the success of generative AI are based on non-mechanistic SSMs treated as input-output mapping modules.
As a result, innovations in likelihood based inference for SSMs could lead to novel techniques for designing and calibrating AI models in ways that are not yet fully understood.
Likelihood based inference for state space models is a large and challenging topic, and this thesis only presents a few innovations related to this problem.
Below, I outline several future projects that build on the work presented in this dissertation.

@@

Additionally, the combination of MPIF and automatic differentiation may facilitate the use of shallow neural networks to be used in conjunction with mechanistic models.
As an example, consider a continuous time compartmental model outlined in Appendix~\ref{subsec:smcs}.
Transition rates between compartments are generally expressed as functions of latent and exogenous variables, denoted $f(\bm{X}(t), \bm{Z}(t))$.
Rather than predetermining the function $f$, automatic differentiation software may enable treating $f$ as a flexible function that needs to be estimated.
A similar idea has been explored previously \citep{noordijk24,dandekar20}, though only in the context of deterministic ODE models, or SDE models with non-mechanistic latent dynamics \citep{lin24}.
In the later approach, a neural network is used to approximate an unknown function that describes the latent process. 
This has the advantage of allowing felxible description of latet states, but it loses the benefits of being mechanistic.
Automatic differentiation for particle filters instead may allow for retaining the general mechanistic framework of a SSM, while permitting extra flexibility for model components where a mechanistic representation is not readily available \citep{dandekar20}.

\subsection*{Applications in Ecology, Agriculture, and Epidemiology}

The MPIF algorithm presented in Chapter~\ref{chpt:mpif} results in an improved capacity to calibrate PanelPOMP models to data via maximum likelihood.
Given the prevalence of panel data and applications of POMP models to univariate time series, this innovation may lead to a number of collaborative research projects in epidemiology.
Iterated filtering has been a poplar approach for modeling infectious disease outbreaks at a single measurement location using nonlinear, overdispersed stochastic models \citep[e.g.,][]{stocks20}. 
The MPIF algorithm facilitates fitting similar models to outbreaks measured at various distinct locations, which can be particularly useful at the onset of disease outbreaks, when there is limited amounts of data available at a single measurement location.

In addition to novel applications in epidemiology, there are several areas of research where I feel that PanelPOMP methodology can lead to novel scientific insights.
Two application areas I feel are particularly promising are fisheries and agriculture.
Fisheries have a rich history of using state space models, though typically inference is performed using Lapacian approximations of the likelihood \citep[e.g.,][]{auger17}.
In addition to comparing whether this approach is a more reliable approximation that those built on sequential Monte Carlo methodologies, studying infectious diseases prevalent in native river and lake species may be uniquely well suited for PanelPOMP methodology: population dynamics in each river or lake are independent from other locations, yet share many similar features that should be estimated simultaneously.
Alternatively, the MPIF methodology may be used to model the nonlinear stochastic dynamics that arise from distinct fish populations competing for the same finite resources \citep{rosenthal22}.
Fish popultations exhibit many of the same obstacles for inference that were outlined by \citet{bjornstad01}, and iterated filtering based algorithms have been successfully used to the same challenges in infectious disease outbreaks.

Livestock populations also exhibit nonlinear, stochastic features of biological systems, and may be well suited for applying PanelPOMP methodology.
Recently published research on livestock management \citep{skolstrup22} used linear Gaussian state space models to estimate the effects of interventions at dairy farms, and methodologies enabling nonlinear models may be a natural next step.
The MPIF algorithm may be used to fit nonlinear SSMs for large collections independent time series data, and could be useful for controling infectious disease outbreaks such as recent avian flu outbreaks \citep{pinotti24}.

### 2025-06-09 chapters/introduction/intro.Rnw
Completed Dissertation Revision

The term state space model (SSM) has a number of synonyms including mechanistic model, partially observed Markov process (POMP) model, and hidden Markov model (HMM) \citep{king15}.
In some literature, these terms have alternatively been used to refer to particular instances of the more general model class \citep[e.g.,][]{glennie23}, resulting in some confusion.
% To avoid ambiguity, unifying notation and terminology is presented in this section.
Throughout the dissertation, I follow the seminal work of \citet{durbin12} and use the term SSM to refer to the most general case.
That is, the observed time series is assumed to be reliant on an unobservable collection of variables at each observation time, without any restrictions on relationships between variables.
The term \emph{mechanistic} is reserved for models where the evolution of latent variables is dictated by equations intended to replicate real-world mechanisms, such as the transmission of infectious diseases among susceptible populations \citep[e.g.,][]{wheeler24}.
In most instances in this thesis and existing literature, the latent variables are treated as a Markov process, in which case the term POMP can be used.
Finally, existing literature often uses the term HMM to refer to POMP models where the latent variables take values in a discrete and finite space \citep{eddy04,doucet01,glennie23,newman23}; this same definition is used here.

@@

popViewport()
@
\caption{\label{fig:pompDiagram}A flow diagram representing an arbitrary POMP model.}
\end{figure}

>>>>

The term state space model (SSM) is often used interchangeably with several synonyms, including mechanistic model, partially observed Markov process (POMP) model, and hidden Markov model (HMM) \citep{king16}.
In some literature, these terms have alternatively been used to refer to particular instances of the more general model class \citep[e.g.,][]{glennie23}, resulting in some confusion.
% To avoid ambiguity, unifying notation and terminology is presented in this section.
Throughout the dissertation, I follow the seminal work of \citet{durbin12} and use the term SSM to refer to the most general case of latent-variable models for time series data.
That is, the observed time series is assumed to be reliant on an unobservable collection of variables at each observation time, without any restrictions on relationships between variables.
In other literature, the term \emph{mechanistic} is used for models that focus on the causality of input-output relationships \citep{baker18} and does not require that the model is a SSM or is used for time series analysis.
Here, the term mechanistic is reserved for SSMs where the evolution of latent variables is dictated by equations intended to replicate real-world mechanisms, such as the transmission of infectious diseases among susceptible populations \citep[e.g.,][]{wheeler24}.
All SSMs considered in this thesis treat the latent variables as a Markov process, in which case the term POMP can be used.
Finally, the term HMM is often used to refer to POMP models where the latent variables take values in a discrete and finite space \citep{eddy04,doucet01,glennie23,newman23}, and the same convention is used here.

@@

popViewport()
@
\caption[A flow diagram representing an arbitrary POMP model.]{\label{fig:pompDiagram}A flow diagram representing an arbitrary POMP model. Figure was built using modified source code from the online course: ``Simulation Based Inference for Epidemiological Dynamics" \citep{king25}.}
\end{figure}

### 2025-06-09 chapters/mpif/ms.Rnw
Completed Dissertation Revision

Contemporary applications of nonlinear mechanistic models for low-dimensional systems are abundant \citep[e.g.,][]{kramer24,he24,newman23}, yet examples of higher-dimensional panel equivalents remain sparse.
In principle, that could be explained by a lack of panel data, or by scientists being uninterested in applying mechanistic models to panel data.
Alternatively, the lack of examples may be indicative that scientists have struggled, largely unsuccessfully, to use existing statistical methodology.
Our work addresses this possibility by proposing a new algorithm that is effective on benchmark problems, theoretically supported, and easy to implement using existing software such as the R package \texttt{panelPomp} \citep{panelpomp}.

>>>>

Contemporary applications of nonlinear mechanistic models for low-dimensional systems are abundant \citep[e.g.,][]{kramer24,he24,newman23}, yet examples of higher-dimensional panel equivalents remain sparse.
In principle, that could be explained by a lack of panel data, or by scientists being uninterested in applying mechanistic models to panel data.
However, previous studies have leveraged simple linear models for panel data and have argued for the importance of fitting models to entire collections of data rather to independent time series \citep[e.g.,][]{wiens14}.
A more plausible explanation is that the lack of examples may be indicative that scientists have struggled, largely unsuccessfully, to use existing statistical methodology.
Our work addresses this possibility by proposing a new algorithm that is effective on benchmark problems, theoretically supported, and easy to implement using existing software such as the R package \texttt{panelPomp} \citep{panelpomp}.
